---
title: "Law of Large Numbers"
author: "Zhentao Shi"
output: pdf_document
---

This script demonstrates the law of large numbers (LLN) along with the underlying assumptions.

Write a function to generate the sample mean given the sample size $n$ and the distribution.
We allow three distributions, namely, $N(0,1)$, $t(2)$ and Cauchy.

```{r}

sample_mean = function( n, distribution ){
  if (distribution == "normal"){ y = rnorm( n ) } 
  else if (distribution == "t2") {y = rt(n, 2) }
  else if (distribution == "cauchy") {y = rcauchy(n) }
  return( mean(y) )
}
```

This function plots the sample mean over the path of geometrically increasing sample size.

```{r}
LLN_plot = function(distribution){
  
  y_bar = rep(0, length(NN) )
  
  for ( i in 1:length(NN)){
    n = NN[i]
    y_bar[i] = sample_mean(n, distribution)
  }
  
  plot(y_bar, type = "l", col = "red", ylab = "mean", xlab = "", lwd = 2, main = distribution)
  abline(h = 0, lty = 2)
  return(y_bar)
}
```

The sample size is chosen as $2^x$, where $x=1:20$. We have the following observations.

* When the distribution is $N(0,1)$, the Chebyshev LLN works. The sample mean converges fast.
* When the distribution is $t(2)$, which has zero mean but infinite variance, the Kolmogorov LLN works. The sample mean still converges, though more slowly than the $N(0,1)$ case.
* The Cauchy distribution has no moment at any order. The sample mean does not converge no matter how large is the sample size.


```{r}
NN = 2^(1:20)
# set.seed(888)

LLN_plot("normal")

LLN_plot("t2")

LLN_plot("cauchy")
```
