---
title: "Panel Data Estimation"
author: "Zhentao Shi"
date: November 22, 2017
output: pdf_document
fontsize: 11pt
---


This document demonstrates panel data estimation methods.

## Dataset


```{r,cache=TRUE}
g0 <- read.csv("http://www.nber.org/nberces/nberces5809/naics5809.csv")
```

The data comes from [NBER-CES Manufacturing Industry Database](http://www.nber.org/nberces/). The data size is about 4M. Downloading would take up to a few minutes if the network is slow. 

The dataset contains annual information of 473 USA industries during 1958 to 2009. To have some idea what a panel data looks like, we display the first 100 rows and 10 columns. 
```{r}
g0[1:80, 1:10]
```

## Estimation

`install.packages("plm")` if you use the package [plm](http://cran.r-project.org/web/packages/plm/) for the first time. An introduction can be found [here](http://cran.r-project.org/web/packages/plm/vignettes/plm.pdf). Load the package.
```{r, message = FALSE}
library(plm)
```

It is very important to explicitly define which column is the cross-sectional dimension and which one is the time dimension.
```{r}
g <- pdata.frame( g0, index = c("naics", "year") )
```
Now we are ready for estimation. Suppose we are interested in, for the purpose of demonstration, a regression with a dependent variable **emp** and explanatory variables **investment** and **capital**. We write down the formula as it will be used repeatedly.
```{r}
equation <- emp~invest+cap
```
### OLS and Pooled OLS
Nothing prevents from running an OLS. 
```{r}
g.ols <- lm(equation, data=g)
summary(g.ols)
```
The OLS coefficient estimates are exactly the same as the pooled OLS. The only difference in the summary is that the later shows the panel structure of the data.


```{r}
g.pool <- plm(equation,data=g,model="pooling")
summary(g.pool)
```

### Random Effect and Fixed Effect
The coefficient estimates differ in the random effect and the fixed effect.
```{r}
g.re <- plm(equation, data=g, model="random")
summary(g.re)
```

```{r}
g.fe <- plm(equation, data=g, model="within")
summary(g.fe)
```
Which model is preferred? The Hausman test favors the fixed-effect model.
```{r}
phtest(g.re, g.fe)
```
